### 2025-04-09 chapters/4-substructure.tex
T-1 day to initial submission, nearly complete draft up to citations and small comments.

\chapter[Removing the Rubble:\\\phantom{Chapter 4:|}Jet Grooming and Cleanup Fish]{Removing the Rubble\\Jet Grooming and Cleanup Fish}
\markboth{\small \textsc{Chapter \thechapter}: \bf Removing the Rubble: Jet Grooming and Cleanup Fish}{}

@@

% -----------------------------------
% Picturebook figure
% -----------------------------------
\vspace{-200pt}
\reusefigure[]{picturebook_substructure}

\begin{itemize}
    \item
We've seen jets as proxies for partons, and discussed how there are difficulties turning this into predictions and comparisons betweeen theory and experiment

    \item
First step towards soft-insensitive jet substructure through jet grooming -- an important suite of tools both experimentally and theoretically which will give us a quick handle on partonic physics even in an experimental context.
\end{itemize}

In this chapter, we introduce \textit{\gls{jet-grooming}}:
%
a powerful and conceptually simple set of tools that facilitate our understanding of jet substructure by removing low-energy pollution from jets.
%
The low-energy phenomena of QCD that produce both soft contamination and \gls{additive-contamination} are notoriously difficult to model, and jet grooming has become an important toolkit used to empower both experimental and theoretical analyses of jet substructure
\cite{Krohn:2009th,Ellis:2009me,Larkoski:2014wba,Dasgupta:2013ihk,Dasgupta:2013via,Tseng:2013dva,Dasgupta:2016ktv,Thaler:2008ju,Thaler:2011gf,Hook:2011cq,Gallicchio:2011xq,Soper:2012pb,Gallicchio:2012ez,CMS-PAS-JME-09-001,CMS-PAS-EXO-09-002,CMS:2013kfa,ATL-PHYS-PUB-2009-081,ATL-PHYS-PUB-2010-008,Cui:2010km,ATLAS-CONF-2011-053,Chatrchyan:2013rla,Larkoski:2013eya,Dasgupta:2012hg,Backovic:2013bga,ATLAS-CONF-2013-084,Komiske:2018vkc,Komiske:2016rsd,Metodiev:2018ftz,Krohn:2012fg,MERINO:2013tta,Bhattacherjee:2016bpy,Macaluso:2018tck,Egan:2017ojy,Kasieczka:2017nvn,Pearkes:2017hku,Butter:2017cot,Catani:1992ua,Dokshitzer:1998kz,Dasgupta:2001sh,Banfi:2004yd,Banfi:2005gj,Ellis:2009su,Banfi:2010pa,Walsh:2011fz,Chien:2012ur,Li:2012bw,Jouttenus:2013hs,Hatta:2013iba,Larkoski:2014tva,Procura:2018zpn,Aaboud:2017aca,Frye:2016aiz, Almeida:2008yp,Larkoski:2017iuy,Larkoski:2017cqq,Thaler:2010tr,Abdesselam:2010pt,Katz:2010mr,Gallicchio:2010dq,Adams:2015hiv,Sirunyan:2017ezt,Moore:2018lsr,FerreiradeLima:2016gcz,Rubin:2010fc,Chatrchyan:2012sn,ATLAS:2019kwg,CMS-PAS-BTV-13-001,CMS-PAS-JME-13-006,Kribs:2009yh,Chen:2010wk,Hackstein:2010wk,Kim:2010uj,Almeida:2011aa,Pandolfi:2012ima,Vernieri:2014wfa,CMS-PAS-HIG-17-007,Procura:2014cba,ATL-PHYS-PUB-2019-027,Aad:2019vyi,ATLAS:2020gwe}.

@@

The conceptual simplicity of jet grooming is accompanied by several practical benefits.
%
Jet grooming can be used to address many sources of low-energy pollution at once, leading to simple analyses that yield robust experimental jet substructure results.
%
Appropriately designed jet grooming strategies can even facilitate analytic calculations of jet substructure observables that may have otherwise involved complicated regions of phase space, or allow the analytic prediction of traditionally mystifying \gls{hadronization} effects \cite{}.

Jet grooming must also be computationally efficient.
%
High-energy particle accelerators collect data from \textit{around one billion collision events every second};%
\footnote{
    This is a concern because jet grooming is used in experimental trigger systems, which filter among these billion events which ones should be stored in the memory of a particle collision database.
}
%
much less importantly, theoretical physicists may get bored if their computers don't spit out an answer immediately.
%
\sam{funny, but you can say something more reasonable here}
%
In order to meet the needs of modern particle physicists, jet grooming algorithms must be as fast as possible.

@@

\remark{jet-discontinuity}{
    All observables depending on a jet definition are technically angularly discontinuous because the angles between particles determine, discontinuously, whether they are reconstructed as within the same jet by the jet algorithm.
    %
    The \gls{kt} and \gls{akt} jet algorithms are soft discontinuous for the same reason.
    %
    Therefore, whenever we use the terms ``soft (angular) continuous'', we implicitly mean ``soft (angular) continuous up to discontinuities associated with jet boundaries/the jet definition itself.''
}

@@

In this chapter, we show that grooming can provide a useful and intuitive way to extract high-energy information from jet substructure.
%
We exposit the physics of traditional, hard-cutoff grooming in \Sec{traditional-grooming} before exploring strategies for continuous grooming -- the \PIRANHA{} paradigm -- in \Sec{piranha}.
%
We emphasize the improved continuity of \PIRANHA{} formally in \Sec{discontinuity} and in practical, phenomenological examples in \Sec{grooming-pheno}.

% TODO: put in table, maybe earlier

@@

\Tab{groomerlist} expresses the infra-red/collinear safety (using the definition of \Reff{Komiske:2020qhg}) and soft/angular continuity properties of the traditional and \PIRANHA{} groomers we discuss in this paper.
%
The soft continuity properties expressed in \Tab{groomerlist} are simple to understand by comparing the hard-cutoff framework of popular groomers to the subtractive framework of \PIRANHA{}.
%
The angular discontinuities expressed in \Tab{groomerlist} are more subtle and are inherited from angular-ordered jet clustering.
%
We note that the Recursive Subtraction algorithms we introduce are discontinuous only in a highly suppressed region of phase space, and we still find it suitable to call P-RS a family of \PIRANHA{} groomers.

Finally, we will achieve the goal of continuous grooming, but not without great toil:
%
as we make grooming procedures which are more and more continuous, they become less and less malleable for theoretical interpretation.
%
Indeed, without some very subtle work, \textit{any} grooming procedure which attempts to utilize the tree-like structure of partonic emissions discussed in the previous chapter will be necessarily discontinuous.
%
Wrestling with the discontinuities of jet grooming and creating continuous methods of extracting partonic information from jets is the continual goal of this section.

@@

In this section, we introduce traditional hard-cutoff techniques for jet grooming, discuss their properties, and understand some analytic features of traditionally-groomed jets.
%
We begin with an extremely simple -- and not impractical -- grooming procedure that is nonetheless similar to more complicated algorithms.
%
We then focus on \glsentrylong{soft-drop} -- an algorithm for jet grooming which is extremely popular due to its simplicity, efficiency, and flexibility.%
\footnote{
    Other traditional grooming algorithms include filtering \cite{}, pruning \cite{}, and some algorithms for boosted object tagging \cite{}, and have similar formal properties to Soft Drop -- including the use of reclustering and the presence of discontinuous behavior.
    % https://arxiv.org/pdf/1901.10342#figure.caption.47
    % https://arxiv.org/pdf/1901.10342#subfigure.6.4.1
    % Resummation not as simple..?
}

Traditional jet grooming algorithms are discontinuous, but we will only briefly touch on the effects of their discontinuous behavior.
%
Therefore, readers familiar with hard-cutoff jet grooming algorithms such as Soft Drop may want to skip to \Sec{discontinuity}, where we discuss the soft discontinuities of the Soft Drop algorithm.
%
In \Sec{grooming-pheno}, we will see the enormous effects of discontinuities in practical applications.

@@

The goal of jet grooming is the efficient removal of low-energy pollution.
%
A simple strategy is to throw away any jet constituents with an energy \(E < E_\text{cut}\), for some constant \(E_\text{cut}\).
%
This is a reasonable first step, and indeed removes low energy radiation from the groomed jet.

@@

\remark{}{
    Since inequalities are discontinuous (for example, \(z > \zcut\) is a discontinuous function from \(z\) to the set \{True, False\}), \glspl{hard-cutoff-groomer} are generically discontinuous as well.
    %
    We say that the constant cutoff algorithm exhibits a \gls{soft-discontinuity} (defined more thoroughly in \Sec{discontinuity}) because the addition of very small amounts of energy can discontinuously change the result of the grooming.
}

@@

\remark{cc-discontinuity}{
    The discontinuity of the \gls{constant-cutoff} grooming algorithm manifests in LL calculations as a \textit{discontinuity in the radiators} of substructure observables, e.g. as in \Eq{ll-cc-mass}.
    %
    Unfortunately, the kink in the radiator presents a complication when attempting to achieve computations beyond LL.
    %
    In particular, the formula of \Eq{multiple-emissions} for the inclusion of multiple-emissions effects on cumulative distributions of additive observables depends on the derivative of the radiator;
    %
    therefore, probability distributions which leverage \Eq{multiple-emissions} depend on the \textit{second derivative} of the radiator.
    %
    Though the radiator turns out to have a well defined derivative, e.g. in \Eq{ll-cc-mass}, it is not second-differentiable, and it is more difficult to account for multiple-emissions effects in \glslink{hard-cutoff-groomer}{hard-cutoff-groomed} jet observables than in ungroomed jet observables.
}

@@

In a constant-cutoff groomed jet, \(\thetacrit\) is a proxy for the first emission in the partonic cascade which satisfies \(z > \zcut\) (similar intuition will hold for \(\thetacrit\) in \gls{soft-drop} groomed jets, as we discuss in the next section).
%
The distribution of \(\zcrit\) is an experimentally accessible probe of the DGLAP splitting functions themselves.\cite{}
%
These observable have a number of practical applications, some of which we will discover ourselves in this chapter, and they are also quite fun to compute on its own.

\begin{exercise}
    Argue that, though \(\thetacrit\) is not an additive observable, we may nonetheless use \glspl{substructure-diagram} and \Eq{ll-groomed-additive} to calculate the cumulative distribution for \(\thetacrit\) at LL.
    %
    In particular, defining \(\Sigma^{\text{(LL)}}_{\thetacrit}\) as the LL probability that no emission with \(z > \zcut\) has \(\theta > \thetacrit\), argue that
    % On the other hand, the probability distribution for \(\theta_{\rm crit}\) alone can be quickly read off from the solution to Example \ref{ex:theta_crit}:
    \begin{align}
        \Sigma^{\text{(LL)}}_{\thetacrit}(\theta_{\rm crit})
        =
        \exp\left[
        ~-~
        \begin{tikzpicture}[
        baseline={([yshift=-.8ex]current bounding box.center)},
        vertex/.style={anchor=base,
        circle,fill=black!25,minimum size=18pt,inner sep=2pt},
        scale=.3]
        \begin{axis}
        [xmin=0, xmax=5,
        ymin=0, ymax=5,
        axis line style = very thick,
        axis y line*=left,
        axis x line*=bottom,
        axis lines = middle,
        ticks=none]
        	\addplot[name path=f,domain=0:2.3,
            style=very thick,blue]
            {2.5};
            \addplot[very thick, samples=50, smooth,domain=0:6,blue, name path=three] coordinates {(2.3,-1)(2.3,2.5)};
            \path[name path=axis]
            (axis cs:0,0) -- (axis cs:2.3,0);
            \addplot [
                thick,
                color=blue,
                fill=blue,
                fill opacity=0.3
            ]
            fill between[
                of=f and axis
            ];
        \end{axis}
        \end{tikzpicture}
        ~~
        \right]
        \approx
        \theta^{-\frac{2 C_R \alpha_{\rm S}}{\pi}\log(z_{\rm cut}^{-1})}
        ,
        \,.
    \end{align}
    What are the bounds on the associated \gls{substructure-diagram}?

@@

    \gls{soft-drop} acts on jets through the following tree-based \gls{declustering} algorithm:
    \begin{enumerate}
        \item
        \label{item:declustering}
        The most recent stage of the C/A clustering is undone, breaking the jet into two sub-jets.

@@

In this work, we use Soft Drop in \vocab{grooming mode}:
%
if Soft Drop reaches a final-state particle that cannot be \glslink{declustering}{de-clustered} as required by Step \ref{item:declustering}, we keep it as the groomed jet.
%
\vocab{Tagging mode} would instead remove it from the jet entirely and leave a groomed event with no particles \cite{Larkoski:2014wba}.

@@

\remark{}{
    The first emission of the angular-ordered jet branching history to pass the grooming procedure will be referred to as the \gls{crit-emission}.
    %
    The ungroomed emissions that occur afterward will be called \glspl{sub-emission}.
    %
    When relevant, we will refer to any branches of the jet history \textit{before} the critical emission as \glspl{precrit-emission}.
}

\gls{soft-drop} \gls{declustering} \sam{Check} is a generalization of the modified Mass Drop Tagger (mMDT) algorithm \cite{Dasgupta:2013ihk}, and reproduces mMDT when \(\beta_{\rm SD} = 0\).
%
Soft Drop has seen success in a wide variety of phenomenological applications, such as the characterization of boosted objects \cite{Thaler:2008ju,Thaler:2011gf,Hook:2011cq,Gallicchio:2011xq,Soper:2012pb,Gallicchio:2012ez,CMS-PAS-JME-09-001,CMS-PAS-EXO-09-002,CMS:2013kfa,ATL-PHYS-PUB-2009-081,ATL-PHYS-PUB-2010-008,ATLAS:2019kwg,Cui:2010km,ATLAS-CONF-2011-053,Chatrchyan:2013rla,Larkoski:2013eya,Dasgupta:2012hg,Backovic:2013bga,ATLAS-CONF-2013-084,Komiske:2018vkc,Komiske:2016rsd,Metodiev:2018ftz,Krohn:2012fg,MERINO:2013tta,Bhattacherjee:2016bpy,Macaluso:2018tck,Egan:2017ojy,Kasieczka:2017nvn,Pearkes:2017hku,Butter:2017cot,Catani:1992ua,Dokshitzer:1998kz,Dasgupta:2001sh,Banfi:2004yd,Banfi:2005gj,Ellis:2009su,Banfi:2010pa,Walsh:2011fz,Chien:2012ur,Li:2012bw,Jouttenus:2013hs,Hatta:2013iba,Larkoski:2014tva,Procura:2018zpn,Aaboud:2017aca,Frye:2016aiz, Almeida:2008yp,Larkoski:2017iuy,Larkoski:2017cqq,Thaler:2010tr,Ellis:2009me,Abdesselam:2010pt,Katz:2010mr,Gallicchio:2010dq,Adams:2015hiv,Sirunyan:2017ezt,Moore:2018lsr,FerreiradeLima:2016gcz,Rubin:2010fc,Chatrchyan:2012sn,CMS-PAS-BTV-13-001,CMS-PAS-JME-13-006,Kribs:2009yh,Chen:2010wk,Hackstein:2010wk,Kim:2010uj,Almeida:2011aa,Pandolfi:2012ima,Vernieri:2014wfa,CMS-PAS-HIG-17-007,Procura:2014cba} and the extraction of parameters of the Standard Model, such as the top quark mass \cite{Hoang:2017kmk,ATLAS:2021urs,Negrini:2022gec} and the strong coupling constant \cite{Marzani:2019evv}, from particle collision data.

@@

    Using the singular pieces of the splitting functions and keeping only leading logarithmic terms in \(\log(m^2/Q^2)\) yields the LL result, for a jet initiated by a parton of color representation \(R\),%
    \footnote{
        The result including non-singular pieces of splitting functions can be found e.g. in \href{https://arxiv.org/pdf/1402.2657\#equation.3.9}{Eq. (3.9)} of \Reff{Larkoski:2014wba} by setting their \(\beta = \betasd\), \(\alpha = 2\).
    }
    %
    \begin{align}
        \label{eq:ll-cc-mass}
        \Sigma^{\rm(LL)}_{m^2}(m^2)
        =
        \begin{cases}
            \exp\le[
                ~-~
                \frac{\alpha_s C_R}{\pi}
                \le(
                    \frac{\betasd \log^2\le(m^2 / Q^2\ri)}
                        {2(2 + \betasd)}
                    +
                    \frac{2 \log(\zcut)\log(m^2/Q^2)}
                        {2 + \beta}
                \ri)
            \ri]
            &
            m^2 < \zcut Q^2
            \,,
            \\
            \exp\le[
                ~-~
                \frac{\alpha_s C_R}{2\pi}
                \,
                \log^2\le(m^2 / Q^2\ri)
            \ri]
            &
            \hspace{-40pt}
            \zcut Q^2 < m^2 < Q^2 / 2
            \,.
        \end{cases}
    \end{align}
    %
    \sam{Fix factors of two!!!}

\remark{}{
        At LL, strong \gls{angular-ordering} dictates that the \glspl{sub-emission} have angles which are much smaller than \(\thetacrit\), \(\theta_\text{sub} \ll \thetacrit\).
    %
    Therefore, in the LL phase space of pQCD, subsequent emissions are \textit{extremely} unlikely to contribute meaningfully to any additive jet observable (which is the sum of contributions which are monotonic in \(\theta\) and vanish as \(\theta \to 0\)).
    %
    Nonetheless, including the effects of the subsequent emissions on additive observables leads to normalizable resummed distributions for observables whose NLO distributions appear to have singular, non-integrable behavior.
    %
    For example, the groomed jet mass distribution we consider above is well-behaved and normalized to unity (\(\Sigma(Q^2/2) = 1\)), but the NLO result, obtained when we expand at \(\mathcal{O}(\alpha_s)\), is singular and not normalizable.
    %
    \sam{Fix factors of two!!!}
}

@@

Though we do not investigate grooming methods that utilize these more restrictive analogs of continuity in this work, the invention and understanding of such groomers offer an intriguing and promising avenue for future research.
%
Instead, the remainder of this work is devoted to introducing hard-cutoff grooming, examining the types of discontinuities that may emerge in jet grooming, and studying the sensitivity of the continuous groomers we introduced in this section to important examples of low-energy pollution that appear in the study of particle collision data.

@@

% \sam{Turn into a problem, as in jet notes?}

>>>>

\chapter[Removing the Rubble:\\\phantom{Chapter 4:|}Jet Grooming and Cleanup Fish]{Removing the Rubble:\hfill\\Jet Grooming and Cleanup Fish}
\markboth{\small \textsc{Chapter \thechapter}: \bf Removing the Rubble: Jet Grooming and Cleanup Fish}{}

@@

% -----------------------------------
% Picturebook figure
% -----------------------------------
\reusefigure[]{picturebook_substructure}

We've now seen jets as proxies for partons, discussed the partonic cascade as a model of jet formation, and given some predictions regarding the jet observables and the flow of energy within jets.
%
We have also seen that there will be difficulties comparing these predictions to experimental results, in the form of low-energy pollution within particle collision events.
%
We now take a first step towards developing soft-insensitive probes of jet substructure, in order to gain information about the flow of energy in the partonic cascade in a more realistic experimental context.

In this chapter, we introduce \textit{\gls{jet-grooming}}:
%
a powerful and conceptually simple set of tools that facilitate our understanding of jet substructure by removing low-energy pollution from jets.
%
The low-energy phenomena of QCD that produce both soft contamination and \gls{additive-contamination} are notoriously difficult to model, and jet grooming has become an important toolkit used to empower both experimental and theoretical analyses of jet substructure
\cite{Krohn:2009th,Ellis:2009me,Larkoski:2014wba,Dasgupta:2013ihk,Dasgupta:2013via,Tseng:2013dva,Dasgupta:2016ktv,Thaler:2008ju,Thaler:2011gf,Hook:2011cq,Gallicchio:2011xq,Soper:2012pb,Gallicchio:2012ez,CMS-PAS-JME-09-001,CMS-PAS-EXO-09-002,CMS:2013kfa,ATL-PHYS-PUB-2009-081,ATL-PHYS-PUB-2010-008,Cui:2010km,ATLAS-CONF-2011-053,Chatrchyan:2013rla,Larkoski:2013eya,Dasgupta:2012hg,Backovic:2013bga,ATLAS-CONF-2013-084,Komiske:2018vkc,Komiske:2016rsd,Metodiev:2018ftz,Krohn:2012fg,MERINO:2013tta,Bhattacherjee:2016bpy,Macaluso:2018tck,Egan:2017ojy,Kasieczka:2017nvn,Pearkes:2017hku,Butter:2017cot,Catani:1992ua,Dokshitzer:1998kz,Dasgupta:2001sh,Banfi:2004yd,Banfi:2005gj,Ellis:2009su,Banfi:2010pa,Walsh:2011fz,Chien:2012ur,Li:2012bw,Jouttenus:2013hs,Hatta:2013iba,Larkoski:2014tva,Procura:2018zpn,Aaboud:2017aca,Frye:2016aiz, Almeida:2008yp,Larkoski:2017iuy,Larkoski:2017cqq,Thaler:2010tr,Abdesselam:2010pt,Katz:2010mr,Gallicchio:2010dq,Adams:2015hiv,Sirunyan:2017ezt,Moore:2018lsr,FerreiradeLima:2016gcz,Rubin:2010fc,Chatrchyan:2012sn,ATLAS:2019kwg,CMS-PAS-BTV-13-001,CMS-PAS-JME-13-006,Kribs:2009yh,Chen:2010wk,Hackstein:2010wk,Kim:2010uj,Almeida:2011aa,Pandolfi:2012ima,Vernieri:2014wfa,CMS-PAS-HIG-17-007,Procura:2014cba,ATL-PHYS-PUB-2019-027,Aad:2019vyi,ATLAS:2020gwe}.
%
Again, I also highly recommend the pedagogical treatment of \Reff{Marzani:2019hun}.

@@

The conceptual simplicity of jet grooming is accompanied by several practical benefits.
%
Jet grooming can be used to address many sources of low-energy pollution at once, leading to simple analyses that yield robust experimental jet substructure results.
%
Appropriately designed jet grooming strategies can even facilitate analytic calculations of jet substructure observables that may have otherwise involved complicated regions of phase space, or allow the analytic prediction of traditionally mystifying \gls{hadronization} effects \cite{Larkoski:2014wba}.

Jet grooming must also be computationally efficient.
%
High-energy particle accelerators collect data from \textit{more than 500 million collision events every second}.%
\footnote{
    This is a concern because jet grooming is used in experimental trigger systems, which filter among these half-a-billion events which ones should be stored in the memory of a particle collision database.
    %
    The given number is from \href{http://lhc-machine-outreach.web.cern.ch/collisions.htm}{\texttt{the CERN website}}, and does not account for fluctuations in the number of collisions.
}
%
In order to meet the needs of modern particle physicists, jet grooming algorithms must be as fast as possible in order to parse through our vast mountains of data.

@@

@@

In order to create continuous methods for extracting high-energy information from jets, we introduce the paradigm of Pileup and Infrared Radiation Annihilation (\PIRANHA{}) for \emph{continuous jet grooming}, introduced by myself and my collaborators in \Reff{Alipour-Fard:2023prj}.
%
The \PIRANHA{} paradigm utilizes lessons from traditional jet grooming together with geometric intuition from optimal transport -- see e.g. \Sec{emd} -- in order to generate more continuous grooming strategies, with better formal and phenomenological behavior.
%
However, though we achieve the goal of continuous grooming through the framework of \PIRANHA{}, we will see that as we make grooming procedures which are more and more continuous, they become less and less malleable for interpretation in perturbative calculations.
%
Indeed, without some very subtle work, \textit{any} grooming procedure which attempts to utilize the tree-like structure of partonic emissions discussed in the previous chapter will be necessarily discontinuous (see \Sec{ang_discont}).

\remark{jet-discontinuity}{
    All observables depending on a jet definition are technically angularly discontinuous because the angles between particles determine, discontinuously, whether a jet algorithm reconstructs them as ling within the same jet.
    %
    Therefore, whenever we use the terms ``soft (angular) continuous'', we implicitly mean ``soft (angular) continuous up to discontinuities associated with jet boundaries/the jet definition itself.''
}

@@

\Tab{groomerlist} expresses the infra-red/collinear safety (using the definition of \Reff{Komiske:2020qhg}) and soft/angular continuity properties of the traditional and \PIRANHA{} groomers we discuss in this paper.
%
The soft continuity properties expressed in \Tab{groomerlist} are simple to understand by comparing the hard-cutoff framework of popular groomers to the subtractive framework of \PIRANHA{}.
%
The angular discontinuities expressed in \Tab{groomerlist} are more subtle and are inherited from angular-ordered jet clustering.
%
We note that the Recursive Subtraction algorithms (P-RS) we introduce are discontinuous only in a highly suppressed region of phase space, and we still find it suitable to call P-RS a family of \PIRANHA{} groomers.

In the remainder of the chapter, we show that grooming can provide a useful and intuitive way to extract high-energy information from jet substructure.
%
We exposit the physics of traditional, hard-cutoff grooming in \Sec{traditional-grooming} before exploring \PIRANHA{} strategies for continuous grooming in \Sec{piranha}.
%
We emphasize the improved continuity of \PIRANHA{} formally in \Sec{discontinuity} and in practical applications in \Sec{grooming-pheno}.

@@

In this section, we introduce traditional hard-cutoff techniques for jet grooming, discuss their properties, and understand some analytic features of traditionally-groomed jets.
%
We begin with an extremely simple -- and not impractical -- grooming procedure that is nonetheless similar to more complicated algorithms.
%
We then focus on \glsentrylong{soft-drop} -- an algorithm for jet grooming which is extremely popular due to its simplicity, efficiency, and flexibility.%
\footnote{
    Other traditional grooming algorithms include filtering, trimming, pruning \cite{Dasgupta:2013ihk}, and some algorithms for boosted object tagging \cite{CMS-PAS-JME-09-001}, and have similar formal properties to Soft Drop -- including the use of reclustering and the presence of discontinuous behavior.
}

Traditional jet grooming algorithms are discontinuous, but we will only briefly touch on the effects of their discontinuous behavior in this section.
%
Therefore, readers familiar with hard-cutoff jet grooming algorithms such as Soft Drop may want to skip to \Sec{discontinuity}, where we discuss the soft discontinuities of the Soft Drop algorithm.
%
In \Sec{grooming-pheno}, we will see the enormous effects of discontinuities in practical applications.

@@

The goal of jet grooming is the efficient removal of low-energy pollution.
%
A simple strategy is to throw away any jet constituents with an energy \(E < E_\text{cut}\), for some constant \(E_\text{cut}\).
%
This is a reasonable first step, and indeed removes low energy radiation to form a resulting groomed jet.

@@

\remark{}{
    Since inequalities are discontinuous (for example, \(z > \zcut\) is a discontinuous function from \(z\) to the set \{True, False\}), \glspl{hard-cutoff-groomer} are generically discontinuous as well.
    %
    We say that the constant cutoff algorithm exhibits a \gls{soft-discontinuity} because the addition of very small amounts of energy can discontinuously change the result of the grooming.
}

@@

\remark{cc-discontinuity}{
    The discontinuity of the \gls{constant-cutoff} grooming algorithm manifests in LL calculations as a \emph{kinds in the radiators} of substructure observables, and associated discontinuities of LL probability distributions (e.g. as in \Eq{ll-cc-mass})
    %
    Unfortunately, the kink in the radiator presents a complication when attempting to achieve computations beyond LL.
    %
    In particular, the formula of \Eq{multiple-emissions} for the inclusion of multiple-emissions effects on cumulative distributions of additive observables depends on the derivative of the radiator;
    %
    therefore, probability distributions which leverage \Eq{multiple-emissions} depend on the \textit{second derivative} of the radiator.
    %
    Though the radiator turns out to have a well defined derivative, e.g. in \Eq{ll-cc-mass}, it is not second-differentiable, and it is more difficult to account for multiple-emissions effects in \glslink{hard-cutoff-groomer}{hard-cutoff-groomed} jet observables than in ungroomed jet observables.
}

@@

In a constant-cutoff groomed jet, \(\thetacrit\) is a proxy for the first emission in the partonic cascade which satisfies \(z > \zcut\) (similar intuition will hold for \(\thetacrit\) in \gls{soft-drop} groomed jets, as we discuss in the next section).
%
The distribution of \(\zcrit\) is an experimentally accessible probe of the DGLAP splitting functions themselves \cite{Larkoski:2015lea,ALargeIonColliderExperiment:2021mqf}.
%
These observable have a number of practical applications, some of which we will discover ourselves in this chapter, and they are also quite fun to compute on its own.

\begin{exercise}
    Argue that, though \(\thetacrit\) is not formally an additive observable, we may nonetheless use \glspl{substructure-diagram} and \Eq{ll-groomed-additive} to calculate the cumulative distribution for \(\thetacrit\) at LL.
    %
    In particular, defining \(\Sigma^{\text{(LL)}}_{\thetacrit}\) as the LL probability that no emission with \(z > \zcut\) has \(\theta > \thetacrit\), argue that
    % On the other hand, the probability distribution for \(\theta_{\rm crit}\) alone can be quickly read off from the solution to Example \ref{ex:theta_crit}:
    \begin{align}
        \Sigma^{\text{(LL)}}_{\thetacrit}(\theta_{\rm crit})
        =
        \exp\left[
        ~-~
        \begin{tikzpicture}[
        baseline={([yshift=-.8ex]current bounding box.center)},
        vertex/.style={anchor=base,
        circle,fill=black!25,minimum size=18pt,inner sep=2pt},
        scale=.3]
        \begin{axis}
        [xmin=0, xmax=5,
        ymin=0, ymax=5,
        axis line style = very thick,
        axis y line*=left,
        axis x line*=bottom,
        axis lines = middle,
        ticks=none]
        	\addplot[name path=f,domain=0:2.3,
            style=very thick,blue]
            {2.5};
            \addplot[very thick, samples=50, smooth,domain=0:6,blue, name path=three] coordinates {(2.3,-1)(2.3,2.5)};
            \path[name path=axis]
            (axis cs:0,0) -- (axis cs:2.3,0);
            \addplot [
                thick,
                color=blue,
                fill=blue,
                fill opacity=0.3
            ]
            fill between[
                of=f and axis
            ];
        \end{axis}
        \end{tikzpicture}
        ~~
        \right]
        \approx
        \theta^{-\frac{2 C_R \alpha_{\rm S}}{\pi}\log(z_{\rm cut}^{-1})}
        ,
        \,.
    \end{align}
    What are the bounds on the associated \gls{substructure-diagram}?

@@

    \gls{soft-drop} acts on jets through the following tree-based \gls{declustering} algorithm:
    \begin{enumerate}
        \item
        \label{item:declustering}
        The most recent stage of the C/A clustering is undone, breaking the jet into two branches/sub-jets.

@@

In the work presented in this thesis, we use Soft Drop in \vocab{grooming mode}:
%
if Soft Drop reaches a final-state particle that cannot be \glslink{declustering}{de-clustered} as required by Step \ref{item:declustering}, we keep it as the groomed jet.
%
\vocab{Tagging mode} would instead remove it from the jet entirely and leave a groomed event with no particles \cite{Larkoski:2014wba}.

@@

\gls{soft-drop} \gls{declustering} is a generalization of the modified Mass Drop Tagger (mMDT) algorithm \cite{Dasgupta:2013ihk}, and reproduces mMDT when \(\beta_{\rm SD} = 0\).
%
Soft Drop has seen success in a wide variety of phenomenological applications, such as the characterization of boosted objects \cite{Thaler:2008ju,Thaler:2011gf,Hook:2011cq,Gallicchio:2011xq,Soper:2012pb,Gallicchio:2012ez,CMS-PAS-JME-09-001,CMS-PAS-EXO-09-002,CMS:2013kfa,ATL-PHYS-PUB-2009-081,ATL-PHYS-PUB-2010-008,ATLAS:2019kwg,Cui:2010km,ATLAS-CONF-2011-053,Chatrchyan:2013rla,Larkoski:2013eya,Dasgupta:2012hg,Backovic:2013bga,ATLAS-CONF-2013-084,Komiske:2018vkc,Komiske:2016rsd,Metodiev:2018ftz,Krohn:2012fg,MERINO:2013tta,Bhattacherjee:2016bpy,Macaluso:2018tck,Egan:2017ojy,Kasieczka:2017nvn,Pearkes:2017hku,Butter:2017cot,Catani:1992ua,Dokshitzer:1998kz,Dasgupta:2001sh,Banfi:2004yd,Banfi:2005gj,Ellis:2009su,Banfi:2010pa,Walsh:2011fz,Chien:2012ur,Li:2012bw,Jouttenus:2013hs,Hatta:2013iba,Larkoski:2014tva,Procura:2018zpn,Aaboud:2017aca,Frye:2016aiz, Almeida:2008yp,Larkoski:2017iuy,Larkoski:2017cqq,Thaler:2010tr,Ellis:2009me,Abdesselam:2010pt,Katz:2010mr,Gallicchio:2010dq,Adams:2015hiv,Sirunyan:2017ezt,Moore:2018lsr,FerreiradeLima:2016gcz,Rubin:2010fc,Chatrchyan:2012sn,CMS-PAS-BTV-13-001,CMS-PAS-JME-13-006,Kribs:2009yh,Chen:2010wk,Hackstein:2010wk,Kim:2010uj,Almeida:2011aa,Pandolfi:2012ima,Vernieri:2014wfa,CMS-PAS-HIG-17-007,Procura:2014cba} and the extraction of parameters of the Standard Model, such as the top quark mass \cite{Hoang:2017kmk,ATLAS:2021urs,Negrini:2022gec} and the strong coupling constant \cite{Marzani:2019evv}, from particle collision data.

@@

    If we use only the singular pieces of the splitting functions, integrate \(z\) from 0 to 1 for simplicity/compactness of final results, and keep only leading logarithmic terms in \(\log(m^2/Q^2)\), we obtain the LL result, for a jet initiated by a parton of color representation \(R\),%
    \footnote{
        The result including additional logarithmic terms in \(\zcut\) and non-singular pieces of splitting functions can be found e.g. in \href{https://arxiv.org/pdf/1402.2657\#equation.3.9}{Eq. (3.9)} of \Reff{Larkoski:2014wba} by setting their \(\beta = \betasd\), \(\alpha = 2\).
    }
    %
    \begin{align}
        \label{eq:ll-sd-mass}
        \Sigma^{\rm(LL)}_{m^2}(m^2)
        =
        \begin{cases}
            \exp\le[
                ~-~
                \frac{\alpha_s C_R}{\pi}
                \le(
                    \frac{\betasd \log^2\le(m^2 / Q^2\ri)}
                        {2(2 + \betasd)}
                    +
                    \frac{2 \log(\zcut)\log(m^2/Q^2)}
                        {2 + \beta}
                \ri)
            \ri]
            &
            m^2 < \zcut Q^2
            \,,
            \\
            \exp\le[
                ~-~
                \frac{\alpha_s C_R}{2\pi}
                \,
                \log^2\le(m^2 / Q^2\ri)
            \ri]
            &
            \hspace{-40pt}
            \zcut Q^2 < m^2 < Q^2 / 2
            \,.
        \end{cases}
    \end{align}
\end{example}

\remark{}{
        At LL, strong \gls{angular-ordering} dictates that the \glspl{sub-emission} have angles which are much smaller than \(\thetacrit\), \(\theta_\text{sub} \ll \thetacrit\).
    %
    Therefore, in the LL phase space of pQCD, subsequent emissions are \textit{extremely} unlikely to contribute meaningfully to any additive jet observable (which is the sum of contributions which are monotonic in \(\theta\) and vanish as \(\theta \to 0\)).
    %
    Nonetheless, including the effects of the subsequent emissions on additive observables leads to normalizable resummed distributions for observables whose NLO distributions appear to have singular, non-integrable behavior.
}

@@

Though we do not investigate grooming methods that utilize these more restrictive analogs of continuity in this thesis, the invention and understanding of such groomers offer an intriguing and promising avenue for future research.
%
Instead, the remainder of this chapter is devoted to introducing hard-cutoff grooming, examining the types of discontinuities that may emerge in jet grooming, and studying the sensitivity of the continuous groomers we introduced in this section to important examples of low-energy pollution that appear in the study of particle collision data.

@@
